Bound within Boundaries: How Well Do Protected Areas Match Movement Corridors of Their Most Mobile Protected Species?

### 2020-09-10 Chapter_0/04_Manuscript/Manuscript.tex
Worked on revisions

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
\noindent 1. Conserving and managing large portions of land to connect wildlife
reserves is increasingly used to maintain and restore connectivity among
wildlife populations. Boundaries of such conservation areas are often determined
based on expert opinion and socio-political constraints, yet the extent to which
they match species' movement corridors is rarely examined. This is mainly due to
a lack of data, particularly on wide-ranging movement behavior such as
dispersal. Nevertheless, empirically assessing the adequacy of protected areas
is key for the implementation of targeted management actions and efficient use
of limited conservation funds.

\noindent 2. Between 2011 and 2019, we collected high-resolution GPS movement
data on 16 dispersing African wild dog (\textit{Lycaon pictus}) coalitions from
a free-ranging population in the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). Spanning five countries and 520'000 km\textsuperscript{2} the
KAZA-TFCA is the world's largest transboundary conservation area and a prime
example for international conservation efforts. We used integrated step
selection analysis to estimate relative selection strenghts\todo{as per reviewer
1} of dispersers and to create a permeability surface for the entire KAZA-TFCA.
We compared landscape permeability across different regions within the KAZA-TFCA
as well as outside its boundaries. Lastly, we calculated least-cost paths and
corridors to verify that major movement routes were adequately encompassed
within the KAZA-TFCA.

\noindent 3. Permeability within the boundaries of the KAZA-TFCA was more than
double compared to areas outside it. Furthermore, we observed a five-fold
permeability difference among the five KAZA-TFCA countries. We further showed
that major movement corridors of wild dogs run within the KAZA-TFCA, although
some minor routes remained outside formally protected areas.

\noindent 4. Differences in permeability were mainly related to different
degrees of human activities across regions, as well as to the presence of
rivers, swamps or open water. The relation between other landscape features and
landscape permeability was less evident.\todo{Editor indicated that the
relationships are correlative and not necessarily causal.}

\noindent 5. \textit{Synthesis and Applications}: In this study, we showed how
pertinent dispersal data of a highly mobile species can be used to empirically
evaluate the adequacy of already-existing or planned protected areas.
Furthermore, observed regional differences in landscape permeability highlight
the need for a coordinated effort towards maintaining or restoring connectivity,
especially where transboundary dispersal occurs.\todo{According to suggestions
by reviewer 2}
\end{abstract}

@@

\section{Introduction}
Connectivity among subpopulations is a crucial pre-requisite for many species to
thrive and persist \citep{Fahrig.2003}. Accordingly, preserving and protecting
movement corridors between wildlife reserves has become a task of utmost
\todo{Reviewer one pointed out an awkward use of the word utmost. I have changed
it according to his suggestion.} importance for conservation management
\citep{Doerr.2011, Rudnick.2012}, resulting in an ever-growing number of large
and often transboundary protected areas. While boundaries of such areas are
often drawn according to expert opinion and socio-political needs, subjective
assessments have revealed deficiencies in the past \citep{Clevenger.2002,
Pullinger.2010}. Thus, an empirical assessment of the adequacy of
already-existing or planned protected areas using pertinent animal movement data
is paramount for targeted use of valuable and scarce conservation funds
\citep{Pullinger.2010}.

In recent years, a growing body of research has used animal relocation data to
identify movement corridors and assess connectivity at large scales (e.g.
\citealp{Chetkiewicz.2006, Squires.2013, Elliot.2014}). Identification of
potential corridors typically relies on the estimation of permeability surfaces,
which return the ease or willingness at which the focal species traverses a
specific landscape \citep{Sawyer.2011}. Such surfaces are created based on a
species' relative selection strengths\todo{reviewer 1}, which can be quantified
using a suite of selection functions \citep{Zeller.2012}. Specifically, relative
selection strenghts \todo{reviewer 1} are estimated by comparing spatial
covariates (e.g. environmental and anthropogenic) at locations visited by the
animal to the same spatial covariates at locations avilable to, yet unused by
\todo{Reviewer 1 pointed out correctly that we should talk about available
locations, rather than randomly selected locations} the animal
\citep{Zeller.2012}. Importantly, selection functions rely on adequate landscape
and relocation data that are representative of the process being studied
\citep{Diniz.2020}. For instance, relocation data collected on dispersing
individuals has been shown to outperform data collected on resident individuals
in the detection of large-scale movement corridors \citep{Elliot.2014,
Diniz.2020}. Nevertheless, dispersal data is inherently difficult to collect and
remains scarce in the connectivity literature \citep{Vasudev.2015}. As such,
most permeability surfaces upon which movement corridors are identified are
created using relocation data collected on resident individuals. While some
species use the same habitats during residence and dispersal
\citep{Fattebert.2015}, others depict vastly different habitat selection
depending on the movement mode \citep{Elliot.2014, Abrahms.2017}. Thus, the use
of data collected during residency can introduce severe biases in estimated
landscape permeability and has likely limited our ability to meaningfully assess
the effectiveness of protected areas in securing connectivity for their
protected species.\todo{Reviewer 2 desired to change from: this introduces - to:
this can introduce}

One initiative that aims at restoring and enhancing connectivity across large
scales is the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA), which
constitutes the world's largest transfrontier conservation area, spanning over
520'000 km\textsuperscript{2} and five countries (\url{www.kavangozambezi.org}).
While the KAZA-TFCA was originally set to facilitate movements of African
elephants (\textit{Loxodonta africana}; \citealp{Tshipa.2017}), it is also key
to the conservation of other wide-ranging species such as African wild dogs
(\textit{Lycaon pictus}; \citealp{Woodroffe.2012, Cozzi.2020}), lions
(\textit{Panthera leo}; \citealp{Elliot.2014, Cushman.2018}), and cheetahs
(\textit{Acinonyx jubatus}; \citealp{Weise.2017}). To date, however, few studies
have attempted to assess the adequacy of the KAZA-TFCA using relevant global
positioning system (GPS) relocation data of its protected species at large
spatial scales \citep{Elliot.2014, Tshipa.2017}. Thus, how well the boundaries
of the KAZA-TFCA reflect natural movement patterns and dispersal corridors of
its most mobile protected species is virtually unknown.

Across the KAZA-TFCA, the African wild dog (\textit{Lycaon pictus}) represents a
highly mobile and endangered flagship species for conservation efforts. Once
widespread across the entire Sub-Saharan continent, wild dogs have been widely
extirpated through human persecution, habitat destruction, and disease outbreaks
\citep{Woodroffe.2012}. As a result, the species has become one of Africa's most
endangered large carnivores, and currently only survives in small, spatially
scattered subpopulations \citep{Woodroffe.2012}. Within these subpopulations,
wild dogs form cooperative breeding packs of up to thirty individuals
\citep{Creel.2002}, whose social structure is strongly governed by the process
of dispersal \citep{McNutt.1996, Behr.2020}. Both males and females disperse
from their natal pack, either alone or in same-sex dispersing coalitions, and
search for unrelated mates and a suitable territory to settle
\citep{McNutt.1996, Cozzi.2020, Behr.2020}. During dispersal, wild dogs can
cover several hundred kilometers \citep{Masenga.2016, Woodroffe.2019,
Cozzi.2020}. Despite the importance of dispersal for the long-term viability of
this species, little empirical information is available on habitat selection and
potential movement barriers during dispersal. The few studies that have
collected dispersal data have shown that dispersers quickly move over large
distances, avoid human-dominated landscapes and areas densely covered by trees,
but prefer proximity to water \citep{Masenga.2016, Woodroffe.2019, Oneill.2020,
Cozzi.2020}.

Here, we collected GPS relocation data on 16 dispersing wild dogs in as many
dispersing coalitions from a free-ranging population in northern Botswana and
analyzed it to assess the adequacy of the KAZA-TFCA in securing connectivity. We
estimated the relative selection strengths\todo{reviewer 1} towards or agains
environmental and anthropogenic landscape features and used the obtained habitat
coefficients to predict a permeability surface spanning the entire KAZA-TFCA. We
then investigated how landscape permeability varies regionally and
internationally and compared permeability within and outside the KAZA-TFCA
boundaries. Finally, we calculated least-cost paths and corridors to identify
major movement routes and to verify that these are successfully covered by the
KAZA-TFCA.

\section{Methods}
\subsection{Study Area}
The study area (centered at -17\degree 13'9''S, 23\degree 56'4''E;
\Cref{StudyArea}a) was outlined by a rectangular bounding box stretching over
1.3 Mio km\textsuperscript{2} and encompassing the entire KAZA-TFCA
(\Cref{StudyArea}b). The KAZA-TFCA lies in the basins of the Okavango and
Zambezi rivers and includes parts of Angola, Botswana, Namibia, Zimbabwe, and
Zambia. With a total area of over 520'000 km\textsuperscript{2} it constitutes
the earth's largest transboundary conservation area and is characterized by
diverse landscapes, including savanna, grassland, and dry or moist woodland
habitats. Rainfall in the study area is seasonal and lasts from November to
March. The KAZA-TFCA also comprises the Okavango Delta, which represents a
highly dynamic hydrological flood-pulsing system \citep{McNutt.1996,
Wolski.2017}. The extent of the flood in the delta greatly changes within and
between years depending on the amount of rain that descends from the catchment
areas in Angola and reaches the distal ends of the delta between July and August
(Figure S4). The flood drastically affects surrounding landscapes, so that
during maximum extent (ca. 12'000 km\textsuperscript{2}) the delta becomes a
patchy conglomerate of swamps, open water, and islands, whereas these structures
run dry when the flood retracts to its minimum extent (ca. 5'000
km\textsuperscript{2}; \citealp{Wolski.2017}). Despite 36 national parks (NPs)
and other protected areas, there is considerable human influence in some regions
of the KAZA-TFCA, mainly originating from farms, high human density, and road
traffic.

@@

\subsection{GPS Relocation Data}
We used a population of free-ranging African wild dogs inhabiting the Okavango
Delta in northern Botswana as a source population for dispersing individuals.
This population has been extensively studied since 1989 \citep{McNutt.1996,
Cozzi.2013, Cozzi.2020, Behr.2020}. Between 2011 and 2019, we systematically
collected GPS relocation data on 16 coalitions of dispersing African wild dogs
(7 female and 9 male coalitions). Candidate dispersing individuals were
identified based on age, number of same‐sex siblings, pack size, and presence of
unrelated individuals of the opposite sex in ther pack citep \citep{McNutt.1996,
Behr.2020}\todo{We originally only pointed to Behr and did not include the
criteria. The editor desired to include those}. Individuals were immobilized
according to protocols described in \cite{Osofsky.1996}, and fitted with
GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic Aerospace GmbH,
Berlin, Germany}) while still with their natal pack. All procedures were
undertaken and supervised by a Botswana-registered wildlife veterinarian.
Collars produced a tag weight of 1.2\% of a wild dog's average body weight,
which is well within published recommendations \citep{Kenward.2000}. Collars
were attached through a leather belt that included a drop off mechanism,
triggered by a slowly decomposing piece of cloth.\todo{Editor desired more
information on GPS collar} During dispersal, GPS collars were programmed to
record GPS relocation data every 4 hours and to regularly transmit them via
Iridium satellite system to a base station.

Because we were interested in dispersal behavior only, we discarded any GPS data
collected while individuals were still with their natal packs and after
settlement in a new territory \citep{Cozzi.2020}. We identified the exact time
of emigration and settlement based on direct field observations and through
visual inspection of the net squared displacement (NSD) metric. NSD quantifies
the squared \todo{Reviewer 1 pointed out correctly that we need to talk about
the \textit{squared} euclidean distance} Euclidean distance of a relocation to a
reference point \citep{Borger.2012}, which in our case was the center of the
dispersing coalition's natal home range. Thus, dispersal was deemed to have
started when a coalition had left its natal home range and continued until the
NSD metric remained stationary, implying that the coalition had successfully
settled (Figure S1). In our analysis, we did not differentiate between male and
female dispersing coalitions, for previous research found little differences
between sexes during dispersal \citep{Woodroffe.2019, Cozzi.2020}.

\subsection{Spatial Covariates}
To investigate relative selection strengths \todo{As per reviewer 1} of
dispersing wild dogs, we used a set of geo-referenced covariates
(\Cref{Covariates}) that we aggregated in the categories \textit{land cover},
\textit{protection status}, and \textit{human influence}. Land cover comprised
of the covariates water cover, distance to water, percentage cover by
shrubs/grassland, and percentage cover by trees. To capture the pulsing behavior
of the Okavango Delta we frequently updated water cover layers and corresponding
distance to water layers. Protection status contained a binary indicator of
whether an area was protected or not. Human influence included covariates
rendering human density, the presence of roads, and the distance to
roads.\todo{Reviewer 2 found the description of the covariates unclear} For each
of these covariates we prepared spatial raster layers from freely available
online services or from remotely sensed satellite imagery. To ensure a
consistent resolution (i.e. cell-size or grain) across covariates, we coarsened
or interpolated all layers to match a resolution of 250m x 250m. For further
details on the preparation of each covariate, see Appendix A.3. We performed
processing and manipulation of data as well as all spatial and statistical
analyses using R, version 3.6.1 \citep{R.2019}.

@@

\subsection{Habitat Selection Model}
We used an integrated step selection function (iSSF; \citealp{Avgar.2016}) to
investigate dispersers' relative selection strentgh \todo{Reviewer 1 suggested
to use relative selection strength, rather than preference} for the
above-mentioned spatial covariates. That is, we paired each realized step (i.e.
the connecting line between two consecutive GPS relocations;
\citealp{Turchin.1998}) with 24 random steps. We generated random steps by
sampling turning angles from a uniform distribution U(\(-\pi, +\pi\)) and step
lengths from a gamma distribution that was fitted using realized steps
\citep{Avgar.2016}. A realized step and its 24 associated random steps formed a
stratum that received a unique identifier. Along each step, we extracted the
above-mentioned covariates (Table S3), standardized extracted values using a
z-score transformation, and checked for correlation using Pearson's Correlation
Coefficient \(r\). None of the covariates were overly correlated (\(|r| > 0.6\);
\citealp{Latham.2011}) and we retained all of them for modeling. Our habitat
selection model then assumed that dispersing wild dogs assigned a selection
score \(w(x)\) of the following exponential form to each realized and random
step \citep{Fortin.2005}:

@@

\noindent The selection score \(w(x)\) of a step depended on its associated
covariates (\(x_1, x_2, ..., x_n\)), as well as on the animal's relative
selection strength \todo{As per reviewer 1} for these covariates (\(\beta_1,
\beta_2, ..., \beta_n\)). To estimate relative selection strength \todo{As per
reviewer 1} (i.e. the \(\beta\)'s) for each covariate, we used mixed effects
conditional logistic regression analysis as suggested by \cite{Muff.2020}. We
implemented their method using the R-package \textit{glmmTMB}
\citep{Mollie.2017} and used dispersing coalition ID to model random slopes. We
also modelled random intercepts with an arbitrary high variance of
10\textsuperscript{6} to make use of the poisson trick (see
\citealp{Muff.2020}). We defined three movement metrics, namely the cosine of
the turning angle (\(cos(ta)\)), the step length (\(sl\)) and the natural
logarithm of the step length (\(log(sl)\)), as core covariates and ran stepwise
forward model selection based on Akaike's Information Criterion (AIC;
\citealp{Burnham.2002}) for all other covariates.\todo{Reviewer 1 asked for more
details in the random effects. I therefore decided to update this section too}
The inclusion of movement metrics served to reduce biases in estimated habitat
selection coefficients \todo{as per revierwer 1} that may have arisen due to
movement behavior \citep{Avgar.2016}. To validate the predictive power of the
most parsimonious habitat selection model, we ran k-fold cross-validation for
case-control studies as described in \cite{Fortin.2009} (details in Appendix
A.5).

\subsection{Permeability Surface}
Using the most parsimonious habitat selection model, we predicted a permeability
surface spanning the entire extent of the KAZA-TFCA. That is, we applied
\ref{EQ1} to our spatial covariates and calculated the selection score \(w(x)\)
for each raster cell. Because our representation of water was dynamic (to
properly render the pulsing behavior of the Okavango Delta) we collapsed all
dynamic water maps into a single static map using areas that were covered by
water in at least 10\% of the cases. Using the resulting static map we also
calculated a layer returning the distance to water. To reduce the influence of
outliers in predicted permeability scores we followed \cite{Squires.2013} and
curtailed predicted scores between the 1\textsuperscript{st} and
99\textsuperscript{th} percentile of their original values. To compare
permeability across different regions, we normalized the permeability surface to
a range between 0 (most impermeable) and 1 (most permeable). We then determined
median permeability within and outside the KAZA-TFCA, within and outside
formally protected areas, and within each of the five KAZA-TFCA countries.

\subsection{Least-Cost Paths and Corridors}
To identify movement corridors of dispersing wild dogs, we specified source
points and calculated factorial least-cost paths (LCPs) as well as factorial
least-cost corridors (LCCs) among them \citep{Elliot.2014}. We generated source
points by overlaying the study area with a regular grid of points spaced at 100
km. We only considered those points that fell within protected areas \(>\) 700
km\textsuperscript{2}, which conforms with home-range requirements of African
wild dogs reported in \cite{Pomilia.2015}. Finally, we defined centroids as
source points for those protected areas \(>\) 700 km\textsuperscript{2} that
were not assigned any source points from the regular grid. In total, we
generated 68 source points, which resulted in 2'278 unique pairwise combinations
and therefore 2'278 unique LCPs and LCCs. We computed factorial LCPs and LCCs
between source points using the R-package \textit{gdistance} (for further
details see Appendix A.7). After computation, we tallied overlapping LCPs and
LCCs, respectively, into single connectivity maps.

>>>>

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
\noindent 1. Conserving and managing large portions of land to connect wildlife
reserves is an increasingly used strategy to maintain and restore connectivity
among wildlife populations. Boundaries of such conservation areas are often
determined based on expert opinion and socio-political constraints, yet the
extent to which they match species' movement corridors is rarely examined. This
is mainly due to a lack of data, particularly on wide-ranging movement behavior
such as dispersal. Nevertheless, empirically assessing the adequacy of protected
areas is key for the implementation of targeted management actions and efficient
use of limited conservation funds.

\noindent 2. Between 2011 and 2019, we collected high-resolution GPS movement
data on 16 dispersing African wild dog (\textit{Lycaon pictus}) coalitions from
a free-ranging population in the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). Spanning five countries and 520'000 km\textsuperscript{2} the
KAZA-TFCA is the world's largest transboundary conservation area and a prime
example for international conservation efforts. We used integrated step
selection analysis to estimate habitat selection of dispersers and to create a
permeability surface for the entire KAZA-TFCA. We compared landscape
permeability across different regions within the KAZA-TFCA as well as outside
its boundaries. Lastly, we calculated least-cost paths and corridors to verify
that major movement routes were adequately encompassed within the KAZA-TFCA.

\noindent 3. Permeability within the boundaries of the KAZA-TFCA was more than
double compared to areas outside it. Furthermore, we observed a five-fold
permeability difference among the five KAZA-TFCA countries. We also showed
that major movement corridors of wild dogs run within the KAZA-TFCA, although
some minor routes remained outside formally protected areas.

\noindent 4. Differences in permeability were mainly related to different
degrees of human activities across regions, and to the presence or absence of
rivers, swamps or open water. The relation between permeability and other
landscape features was less evident.\todo{Editor indicated that the
relationships are correlative and not necessarily causal.}

\noindent 5. \textit{Synthesis and Applications}: In this study, we showed how
pertinent dispersal data of a highly mobile species can be used to empirically
evaluate the adequacy of already-existing or planned protected areas.
Furthermore, we observed regional differences in landscape permeability that
highlight the need for a coordinated effort towards maintaining or restoring
connectivity, especially where transboundary dispersal occurs.\todo{According to
suggestions of reviewer 2}
\end{abstract}

@@

\section{Introduction}
Connectivity among subpopulations is a crucial pre-requisite for many species to
thrive and persist \citep{Fahrig.2003}. Accordingly, preserving and protecting
movement corridors between wildlife reserves has become a task of utmost
\todo{Reviewer one pointed out an awkward use of the word utmost. I have changed
it according to his suggestion.} importance \citep{Doerr.2011, Rudnick.2012},
resulting in an ever-growing number of large and often transboundary protected
areas. While boundaries of such areas are often drawn according to expert
opinion and socio-political needs, subjective assessments have revealed
deficiencies in the past \citep{Clevenger.2002, Pullinger.2010}. Thus, an
empirical evaluation of the adequacy of already-existing or planned protected
areas using pertinent animal movement data is paramount for targeted use of
valuable and scarce conservation funds \citep{Pullinger.2010}.

In recent years, a growing body of research has used animal relocation data to
identify movement corridors and assess connectivity at large scales (e.g.
\citealp{Chetkiewicz.2006, Squires.2013, Elliot.2014}). The use of relocation
data has shifted the focus from a structural to a functional view on
connectivity, taking into account how species respond to environmental features
\citep{Taylor.2006, Fattebert.2015}.\todo{As per reviewer 2} Identification of
potential movement corridors typically relies on the estimation of permeability
surfaces, which return the ease or willingness at which the focal species
traverses a specific landscape \citep{Sawyer.2011}. Such surfaces are created
based on species' relative selection strengths \citep{Avgar.2017}\todo{reviewer
1}, which can be quantified using a suite of selection functions
\citep{Zeller.2012}. Specifically, selection strengths \todo{reviewer 1} are
estimated by comparing spatial covariates (e.g. environmental and anthropogenic)
at locations visited by the animal to the same spatial covariates at locations
avilable to the animal \todo{Reviewer 1 pointed out correctly that we should
talk about available locations, rather than randomly selected locations}
\citep{Zeller.2012}. Importantly, selection functions rely on adequate landscape
and relocation data that are representative of the process being studied
\citep{Diniz.2020}. Altough selection during residence and dispersal may
coincide \citep{Fattebert.2015},\todo{second reviewer pointed out that space use
can be alike} it appears that relocation data collected on dispersing
individuals outperforms data collected on resident individuals in the detection
of large-scale movement corridors \citep{Elliot.2014, Diniz.2020}. Nevertheless,
dispersal data is inherently difficult to collect and remains scarce in the
connectivity literature \citep{Vasudev.2015}. As such, most permeability
surfaces upon which movement corridors are identified are created using
relocation data collected on resident individuals. This has likely limited our
ability to meaningfully assess the effectiveness of protected areas in securing
functional connectivity for their protected species.\todo{Reviewer 2 desired to
change from: this introduces - to: this can introduce}

One initiative that aims at restoring and enhancing connectivity across large
scales is the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA), which
constitutes the world's largest transfrontier conservation area, spanning over
520'000 km\textsuperscript{2} and five countries (\url{www.kavangozambezi.org}).
While the KAZA-TFCA was originally set to facilitate movements of African
elephants (\textit{Loxodonta africana}; \citealp{Tshipa.2017}), it is also key
to the conservation of other wide-ranging species such as African wild dogs
(\textit{Lycaon pictus}; \citealp{Woodroffe.2012, Cozzi.2020}), lions
(\textit{Panthera leo}; \citealp{Elliot.2014, Cushman.2018}), and cheetahs
(\textit{Acinonyx jubatus}; \citealp{Weise.2017}). To date, however, few studies
have attempted to assess the adequacy of the KAZA-TFCA using relevant global
positioning system (GPS) relocation data of its protected species at large
spatial scales \citep{Elliot.2014, Tshipa.2017}.\todo{as per editor} Thus, how
well the boundaries of the KAZA-TFCA reflect natural movement patterns and
dispersal corridors of its most mobile protected species is virtually unknown.

Across the KAZA-TFCA, the African wild dog (\textit{Lycaon pictus}) represents a
highly mobile and endangered flagship species for conservation efforts. Once
widespread across the entire Sub-Saharan continent, wild dogs have been widely
extirpated through human persecution, habitat destruction, and disease outbreaks
\citep{Woodroffe.2012}. As a result, the species has become one of Africa's most
endangered large carnivores, and currently only survives in small, spatially
scattered subpopulations \citep{Woodroffe.2012}. Within these subpopulations,
wild dogs form cooperative breeding packs of up to thirty individuals
\citep{Creel.2002}, whose social structure is strongly governed by the process
of dispersal \citep{McNutt.1996, Behr.2020}. Both males and females disperse
from their natal pack, either alone or in same-sex dispersing coalitions, and
search for unrelated mates and a suitable territory to settle
\citep{McNutt.1996, Cozzi.2020, Behr.2020}. During dispersal, wild dogs can
cover several hundred kilometers and cross international borders
\citep{Masenga.2016, Woodroffe.2019, Cozzi.2020}. Despite the importance of
dispersal for the long-term viability of this species, little empirical
information is available on habitat selection and potential movement barriers
during dispersal. The few studies that have collected dispersal data have shown
that dispersers quickly move over large distances, avoid human-dominated
landscapes and areas densely covered by trees, but prefer proximity to water
\citep{Masenga.2016, Woodroffe.2019, Oneill.2020, Cozzi.2020}.

Here, we collected and analyzed GPS relocation data on 16 dispersing wild dogs
in as many dispersing coalitions from a free-ranging population in northern
Botswana to assess the adequacy of the KAZA-TFCA in securing connectivity. We
estimated relative selection strengths\todo{reviewer 1} towards environmental
and anthropogenic landscape features, and used the obtained coefficients to
predict a permeability surface spanning the entire KAZA-TFCA. We then
investigated how landscape permeability varied regionally and internationally,
and we compared permeability within and outside the KAZA-TFCA boundaries.
Finally, we calculated least-cost paths and corridors to identify major movement
routes and to verify that these were successfully covered by the KAZA-TFCA.

\section{Methods}
\subsection{Study Area}
The study area (centered at -17\degree 13'9''S, 23\degree 56'4''E;
\Cref{StudyArea}a) was outlined by a rectangular bounding box stretching over
1.3 Mio km\textsuperscript{2} and encompassing the entire KAZA-TFCA
(\Cref{StudyArea}b). The KAZA-TFCA lies in the basins of the Okavango and
Zambezi rivers and includes parts of Angola, Botswana, Namibia, Zimbabwe, and
Zambia. With a total area of over 520'000 km\textsuperscript{2} it constitutes
the earth's largest transboundary conservation area and is characterized by
diverse landscapes, including savanna, grassland, and dry or moist woodland
habitats. Rainfall in the study area is seasonal and lasts from November to
March. The KAZA-TFCA also comprises the Okavango Delta, which represents a
highly dynamic hydrological flood-pulsing system \citep{McNutt.1996,
Wolski.2017}. The extent of the flood in the delta greatly changes within and
between years, depending on the amount of rain that descends from the catchment
areas in Angola and reaches the distal ends of the delta between July and August
(Figure S4). The flood drastically affects surrounding landscapes, so that
during maximum extent (ca. 12'000 km\textsuperscript{2}) the delta becomes a
patchy conglomerate of swamps, open water, and islands, whereas these structures
run dry when the flood retracts to its minimum extent (ca. 5'000
km\textsuperscript{2}; \citealp{Wolski.2017}). Despite 36 national parks (NPs)
and other protected areas, there is considerable human influence in some regions
of the KAZA-TFCA, mainly originating from farms, human density, and road
traffic.

@@

\subsection{GPS Relocation Data}
We used a population of free-ranging African wild dogs inhabiting the Okavango
Delta in northern Botswana as a source population for dispersing individuals.
This population has been extensively studied since 1989 \citep{McNutt.1996,
Cozzi.2013, Cozzi.2020, Behr.2020}. Between 2011 and 2019, we systematically
collected GPS relocation data on 16 coalitions of dispersing African wild dogs
(7 female and 9 male coalitions). Candidate dispersing individuals were
identified based on age, number of same‐sex siblings, pack size, and presence of
unrelated individuals of the opposite sex in their pack \citep{McNutt.1996,
Behr.2020}\todo{The editor desired to include criteria}. Selected individuals
were immobilized according to protocols described in \cite{Osofsky.1996}, and
fitted with GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic
Aerospace GmbH, Berlin, Germany}) while still with their natal pack. All
required procedures were undertaken and supervised by a Botswana-registered
wildlife veterinarian. Collars produced a tag weight of 300 gramms, accounting
for 1.2\% of a wild dog's average body weight. They were attached using leather
belts that included a drop off mechanism, triggered by a decomposing piece of
cloth.\todo{Editor desired more information on GPS collar} During dispersal, GPS
collars were programmed to record a GPS relocation every 4 hours and to
regularly transmit data via Iridium satellite system to a base station.

Because we were interested in dispersal behavior only, we discarded any GPS data
collected while individuals were still with their natal packs and after
settlement in a new territory \citep{Cozzi.2020}. We identified the exact time
of emigration and settlement based on direct field observations and through
visual inspection of the net squared displacement (NSD) metric. NSD quantifies
the squared \todo{Reviewer 1 pointed out correctly that we need to talk about
the \textit{squared} euclidean distance} Euclidean distance of a relocation to a
reference point \citep{Borger.2012}, which in our case was the center of the
dispersing coalition's natal home range. Thus, dispersal was deemed to have
started when a coalition had left its natal home range and continued until the
NSD metric remained stationary, indicating that the coalition had successfully
settled (Figure S1). In our analysis, we did not differentiate between male and
female dispersing coalitions, for previous research found little differences
between sexes during dispersal \citep{Woodroffe.2019, Cozzi.2020}.

\subsection{Spatial Covariates}
To investigate relative selection strengths \todo{As per reviewer 1} of
dispersing wild dogs, we used a set of geo-referenced covariates
(\Cref{Covariates}) that we aggregated in the categories \textit{land cover},
\textit{protection status}, and \textit{human influence}. Land cover comprised
the covariates water cover, distance to water, percentage cover by
shrubs/grassland, and percentage cover by trees. To capture the pulsing behavior
of the Okavango Delta, we frequently updated layers for water cover and
corresponding layers depicting distance to water. Protection status contained a
binary indicator of whether an area was protected or not. Human influence
included covariates rendering human density, the presence of roads, and the
distance to roads.\todo{Reviewer 2 found the description of the covariates
unclear} We prepared all covariates as spatial raster layers from freely
available online services and from remotely sensed satellite imagery. To ensure
a consistent resolution (i.e. cell-size or grain) across covariates, we
coarsened or interpolated all layers to a resolution of 250m x 250m. For further
details on the preparation of each covariate and their sources, see Appendix
A.3. We performed processing and manipulation of data as well as all spatial and
statistical analyses using R, version 3.6.1 \citep{R.2019}.

@@

\subsection{Habitat Selection Model}
We used an integrated step selection function (iSSF; \citealp{Avgar.2016}) to
investigate dispersers' relative selection strentghs towards \todo{Reviewer 1
suggested to use relative selection strength, rather than preference} the
above-mentioned spatial covariates. That is, we paired each realized step (i.e.
the connecting line between two consecutive GPS relocations;
\citealp{Turchin.1998}) with 24 random steps that were generated by sampling
turning angles from a uniform distribution U(\(-\pi, +\pi\)) and step lengths
from a gamma distribution fitted to realized steps \citep{Avgar.2016}. A
realized step and its 24 associated random steps formed a stratum and received a
unique identifier. Along each step, we extracted the above-mentioned covariates
(Table S3), standardized extracted values using a z-score transformation, and
checked for correlation using Pearson's Correlation Coefficient \(r\). None of
the covariates were overly correlated (\(|r| > 0.6\); \citealp{Latham.2011}) and
we retained all of them for modeling. Our habitat selection model then assumed
that dispersing wild dogs assigned a selection score \(w(x)\) of the following
exponential form to each realized and random step \citep{Fortin.2005}:

@@

\noindent The selection score \(w(x)\) of a step depended on its associated
covariates (\(x_1, x_2, ..., x_n\)), as well as on the animal's relative
selection strengths \todo{As per reviewer 1} towards these covariates
(\(\beta_1, \beta_2, ..., \beta_n\)). To estimate relative selection strengths
\todo{As per reviewer 1} (i.e. the \(\beta\)'s) for each covariate, we used
mixed effects conditional logistic regression analysis as suggested by
\cite{Muff.2020}. We implemented their method using the R-package
\textit{glmmTMB} \citep{Mollie.2017} and used dispersing coalition ID to model
random slopes. We also modelled random intercepts with an arbitrary high
variance of 10\textsuperscript{6} to make use of the poisson trick (see
\citealp{Muff.2020}). We defined three movement metrics, namely the cosine of
the turning angle (\(cos(ta)\)), the step length (\(sl\)) and the natural
logarithm of the step length (\(log(sl)\)), as core covariates and ran stepwise
forward model selection based on Akaike's Information Criterion (AIC;
\citealp{Burnham.2002}) for all other covariates.\todo{Reviewer 1 asked for more
details in the random effects. I therefore decided to update this section too}
The inclusion of movement metrics served to reduce biases in estimated habitat
selection coefficients \todo{as per revierwer 1} that may have arisen due to
movement behavior \citep{Avgar.2016}. To validate the predictive power of the
most parsimonious habitat selection model, we ran k-fold cross-validation for
case-control studies as described in \cite{Fortin.2009} (details in Appendix
A.5).

\subsection{Permeability Surface}
Using the most parsimonious habitat selection model, we predicted a permeability
surface spanning the entire extent of the KAZA-TFCA. That is, we applied
\ref{EQ1} to our spatial covariates and calculated the selection score \(w(x)\)
for each raster cell. Because our representation of water was dynamic (to
properly render the pulsing behavior of the Okavango Delta) we collapsed all
dynamic water maps into a single static map using areas that were covered by
water in at least 10\% of the cases. Using the resulting static map we also
calculated a layer returning the distance to water. Because the extent of the
delta covers merely 6\% of the entire KAZA-TFCA, the use of static maps should
introduce little bias in predicted permeability. To reduce the influence of
outliers in predicted permeability scores, we followed \cite{Squires.2013} and
curtailed predicted scores between the 1\textsuperscript{st} and
99\textsuperscript{th} percentile of their original values. To compare
permeability across different regions, we normalized the permeability surface to
a range between 0 (most impermeable) and 1 (most permeable), and we determined
median permeability within and outside the KAZA-TFCA, within and outside
formally protected areas, and within each of the five KAZA-TFCA countries.

\subsection{Least-Cost Paths and Corridors}
To identify movement corridors of dispersing wild dogs, we specified source
points and calculated factorial least-cost paths (LCPs) as well as factorial
least-cost corridors (LCCs) among them \citep{Elliot.2014}. We generated source
points by overlaying the study area with a regular grid of points spaced at 100
km. We only considered those points that fell within protected areas \(>\) 700
km\textsuperscript{2}, which conforms with home-range requirements of African
wild dogs \citep{Pomilia.2015}. Finally, we defined centroids as source points
for those protected areas \(>\) 700 km\textsuperscript{2} that were not assigned
any source points from the regular grid. Because wild dogs residing outside of
protected areas are rare and likely unviable \citep{VanDerMeer.2014}, we
consider our selection of source points within protected areas to be
appropriate. In total, we generated 68 source points, which resulted in 2'278
unique pairwise combinations and therefore 2'278 unique LCPs and LCCs. We
computed factorial LCPs and LCCs between source points using the R-package
\textit{gdistance} (further details in Appendix A.7). After computation, we
tallied overlapping LCPs and LCCs, respectively, into single connectivity maps.

\section{Results}
\subsection{Dispersal Events}
In total, we collected 4'169 GPS relocations during dispersal (Figure S2 \&
Table S1), resulting in an average of 261 (\(SD\) = 207) locations per
dispersing coalition. Coalitions on average dispersed for 48 days (\(SD\) = 44),
covered a mean Euclidean distance of 54 km (\(SD\) = 71) and a cumulative
distance of 597 km (\(SD\) = 508). One female coalition dispersed far east into
the Hwange Hwange National Park and covered a cumulative distance of over 360 km
in under 9 days.\todo{Editor desired a quick summary of the dispersal data.}
